Understanding and Verifying the Shape of Stencil Computations

### 2017-03-26 oopsla17/lattice-model.tex
Add few macros, add some goals for inference procedure

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{indentfirst}
\usepackage{cleveref}

\theoremstyle{definition}
\newtheorem{defn}{Definition}

@@

\newcommand{\eg}{\emph{e.g.}}

\newcommand{\zinf}{\textnormal{$\mathbb{Z}_\infty$}}
\newcommand{\interp}[1]{\llbracket#1\rrbracket}
\newcommand{\interv}[3]{\textnormal{$\lbrack#1,#2\rbrack^{#3}$}}
\newcommand{\vecgen}[3]{\textnormal{$\textit{vec-gen}_{#1}(#2,#3)$}}
\newcommand{\textcap}[1]{\texttt{\textit{#1}}}

@@

\begin{lem}{}
  We have the following dual identities for \zinf{} intervals:
%
  \begin{align*}
    \interv{a}{b}{c} \cap \interv{d}{e}{f} & =
      \interv{\max \{a,d\}}{\min \{b,e\}}{c \wedge f} \\
    \interv{a}{b}{c} \cup \interv{d}{e}{f} & =
      \interv{\min \{a,d\}}{\max \{b,e\}}{c \vee f}
  \end{align*}\label{lem:zinf-identities}
\end{lem}
%
\begin{proof}
  We give the proof of the first identity and the second one is similar.
  \begin{align*}
    \interv{a}{b}{c} \cap \interv{d}{e}{f} = &
      \; \{ n \;|\; a \leq n \leq b \wedge (\neg c \implies n \neq 0) \}
      \;\cap \\
      & \; \{ n \;|\; d \leq n \leq e \wedge (\neg f \implies n \neq 0) \}
      \\
    = & \; \{ n \;|\; \max \{a,d\} \leq n \leq \min \{b,e\} \wedge (\neg c
      \vee \neg f \implies n \neq 0) \} \\
    = & \; \interv{\max{\{a,d\}}}{\min{\{b,e\}}}{\neg c \vee \neg f}
  \end{align*}
\end{proof}

@@

\begin{lem}{}
  Intersection distributes over offsets vectors.
%
  \begin{equation*}
    (\textit{vint}_1, \dots, \textit{vint}_N) \;\cap\;
      (\textit{vint'}_1, \dots, \textit{vint'}_N)
      =
    (\textit{vint}_1 \;\cap\; \textit{vint'}_1, \dots,
     \textit{vint}_N \;\cap\; \textit{vint'}_N)
  \end{equation*}\label{lem:vector-intersect}
\end{lem}

\begin{lem}{}
  Let $v$ and $v'$ be two $N$ dimensional offsets vectors such that $v_i$ is
  equal to $v'_i$ for all $1 \leq i \leq N$ except $k$, then we have the
  following:
%
  \begin{equation*}
    (v_1,\dotsc, v_k,\dotsc, v_N) \; \cup \; (v'_1,\dotsc, v'_k, \dotsc, v_N)
    =
    (v_1,\dotsc, v_k \cup v'_k,\dotsc, v_N)
  \end{equation*}\label{lem:vector-union}
\end{lem}
%
\begin{proof}
  \begin{align*}
    v \; \cup \; v' &
    = \{(x_1,\dotsc,x_n) \mid
          \bigwedge_{1 \leq i \leq N } x_i \in v_i \}
      \;\cup\;
      \{(x_1,\dotsc,x_n) \mid
          \bigwedge_{1 \leq i \leq N } x_i \in v'_i \} \\
    & = \{(x_1,\dotsc,x_n) \mid
          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}} x_i \in v_i
              \wedge x_k \in v_k \vee
          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}} x_i \in v'_i
            \wedge x_k \in v'_k
        \} \\
    & = \{(x_1,\dotsc,x_n) \mid
          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}} x_i \in v_i
          \wedge x_k \in v_k \cup v'_k
        \} \\
    & = (v_1,\dotsc, v_k \cup v'_k,\dotsc, v_N)
  \end{align*}
\end{proof}

@@

\begin{lem}{}
  Intersection distributes $N$ dimensional vector intervals:
%
  \begin{gather*}
    (\interv{a_1}{b_1}{c_1}, \dots, \interv{a_N}{b_N}{c_N}) \; \cap \;
      (\interv{a'_1}{b'_1}{c'_1}, \dots, \interv{a'_N}{b'_N}{c'_N}) \\
    = \\
    (\interv{\max \{a_1,a'_1\}}{\min \{b_1,b'_1\}}{\neg c_1 \vee \neg c'_1},
      \dots, \interv{\max \{a_N,a'_N\}}{\min \{b_N,b'_N\}}{\neg c_N \vee \neg c'_N})
  \end{gather*}
\end{lem}
%
\begin{proof}
  This immediately follows from \Cref{lem:vector-intersect} and
  \Cref{lem:zinf-identities}.
\end{proof}

@@

We give some basic interpretations: $\interp{\texttt{n}} = n$ where $n$ is in
$\mathbb{Z}$, $\interp{\texttt{pointed}} = \llbracket \epsilon \rrbracket =
\textsf{true}$, $\interp{\texttt{nonpointed}} = \textsf{false}$.
Typographically, cursive font is used within specifications to indicate a
partial textual capture.

@@

First, we give meaning to region constants (introduced in \Cref{}).
%
\begin{align*}
  \interp{\texttt{pointed(dim=\textcap{i})}}_N & =
    \vecgen{N}{\interp{\textcap{i}}}{\interv{0}{0}{\textsf{true}}}\\
%
  \interp{\texttt{centered(dim=\textcap{i}, depth=\textcap{k}, \textcap{p})}}_N
    & = \vecgen{N}
           {\interp{\textcap{i}}}
           {\interv{-\interp{\textcap{k}}}
                   {\interp{\textcap{k}}}
                   {\interp{\textcap{p}}}
           } \\
%
  \interp{\texttt{forward(dim=\textcap{i}, depth=\textcap{k}, \textcap{p})}}_N
    & = \vecgen{N}
           {\interp{\textcap{i}}}
           {\interv{0}
                   {\interp{\textcap{k}}}
                   {\interp{\textcap{p}}}
           } \\
%
  \interp{\texttt{backward(dim=\textcap{i}, depth=\textcap{k}, \textcap{p})}}_N
    & = \vecgen{N}
           {\interp{\textcap{i}}}
           {\interv{-\interp{\textcap{k}}}
                   {0}
                   {\interp{\textcap{p}}}
           }
\end{align*}

@@

We know introduce the equational and approximation theory of stencil
specifications based on $\equiv$ and $\prec$.

@@

\begin{description}
  \item[Basic] \texttt{*} and \texttt{+} are both idempotent, commutative, and
    associative;
%
  \item[Dist] \texttt{*} distributed over \texttt{+} and dually
    \texttt{+} distributes over \texttt{*}, meaning if \textcap{R}, \textcap{S},
    and \textcap{T} are regions, then we have the following equivalences:
%
    \begin{align*}
      \texttt{\textcap{R}*(\textcap{S}+\textcap{T})} & \equiv
        \texttt{(\textcap{R}*\textcap{S})+(\textcap{R}*\textcap{T})} &
      \texttt{\textcap{R}+(\textcap{S}*\textcap{T})} & \equiv
        \texttt{(\textcap{R}+\textcap{S})*(\textcap{R}+\textcap{T})}
    \end{align*}
%
  \item[Overlapping pointed] If \textcap{R} is one of \texttt{forward},
    \texttt{backward}, or \texttt{centered}, then we have
    $\texttt{\textcap{R}(depth=\textcap{k},dim=\textcap{n},\_) +
    pointed(dim=\textcap{n})} \equiv
    \texttt{\textcap{R}(depth=\textcap{k},dim=\textcap{n})}$.
%
  \item[Centered] The region constants \texttt{forward} and \texttt{backward}
    are two halves of \texttt{centered} specifications:
%
    \begin{gather*}
      \texttt{centered(depth=\textcap{k},dim=\textit{n},\textcap{p1})} \\
      \equiv \\
      \texttt{forward(depth=\textcap{k},dim=\textit{n},\textcap{p2}) +
        backward(depth=\textcap{k},dim=\textit{n},\textcap{p3})}
    \end{gather*}
%
    Here \textcap{p1} is \texttt{nonpointed} if both \textcap{p2} and
    \textcap{p3} are \texttt{nonpointed} and \texttt{pointed} otherwise.
%
  \item[Point] Let \textcap{R} and \textcap{S} be region constants with two
    variants a superscript \textit{f} to mean nonpointed and \textit{t} to mean
    pointed and \textcap{dr} \& \textcap{ds} be the dimensions \textcap{R} and
    \textcap{S} acts on. Then we have the following dual equivalences:
%
    \begin{align*}
      (\textcap{R}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{f}})
      \;\texttt{+}\;
      (\textcap{R}^{\textit{f}} \texttt{*} \textcap{S}^{\textit{t}})
      \;\texttt{+}\;
      \texttt{p(dim=\textcap{dr})*p(dim=\textcap{ds})} & =
        \textcap{R}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{t}} \\
%
      (\textcap{R}^{\textit{t}} \texttt{+} \textcap{S}^{\textit{f}}) \texttt{*}
      (\textcap{R}^{\textit{f}} \texttt{+} \textcap{S}^{\textit{t}}) & =
        \textcap{R}^{\textit{f}} \texttt{+} \textcap{S}^{\textit{f}}
        \;\texttt{+}\;
        \texttt{p(dim=\textcap{dr})*p(dim=\textcap{ds})}
    \end{align*}
\end{description}

@@

\subsection{Approximations}

@@

>>>>

% Essential packages
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{stmaryrd}
\usepackage{indentfirst}
\usepackage{cleveref}
\usepackage{xifthen}

% Non-essential
\usepackage[usenames,dvipsnames,svgnames,table]{xcolor}

% Essential declarations
\theoremstyle{definition}
\newtheorem{defn}{Definition}

@@

% Auxillary declaration
\newcommand{\eg}{\emph{e.g.}}

% Essential macros
\newcommand{\zinf}{\textnormal{$\mathbb{Z}_\infty$}}
\newcommand{\interp}[1]{\llbracket {#1} \rrbracket}
\newcommand{\interv}[3]{\textnormal{$\lbrack#1,#2\rbrack^{#3}$}}
\newcommand{\vecgen}[3]{\textnormal{$\textit{vec-gen}_{#1}(#2,#3)$}}
\newcommand{\textcap}[1]{\texttt{\textit{#1}}}

% Stencil macro
%
% Optional argument: s when we want the short form, empty otherwise.
%
% Mandatory arguments:
% 1st: c/b/f/p for centered, backward, forward, pointed
% 2st: dimension
% 3st: depth
% 4st: if pointed/nonpointed needs to be specified leave it blank if omitted.
%
% Example: \stencil[s]{f}{1}{2}{} = fwd(dim=1,depth=2)
% Example: \stencil{b}{1}{2}{pointed} = backward(dim=1,depth=2,pointed)
\newcommand{\stencil}[5][]{%
  \ifthenelse{\equal{#2}{f}}{%
    \ifthenelse{\equal{#1}{s}}{%
      \texttt{fwd}%
    }{\ifthenelse{\isempty{#1}}{%
        \texttt{forward}%
      }{INVALID OPTIONAL}%
    }%
  }{\ifthenelse{\equal{#2}{b}}{%
      \ifthenelse{\equal{#1}{s}}{%
          \texttt{bwd}%
      }{\ifthenelse{\isempty{#1}}{%
          \texttt{backward}%
        }{INVALID OPTIONAL}%
      }%
  }{\ifthenelse{\equal{#2}{c}}{%
      \ifthenelse{\equal{#1}{s}}{%
          \texttt{ctd}%
      }{\ifthenelse{\isempty{#1}}{%
          \texttt{centered}%
        }{INVALID OPTIONAL}%
      }%
  }{\ifthenelse{\equal{#2}{p}}{%
      \ifthenelse{\equal{#1}{s}}{%
          \texttt{ptd}%
      }{\ifthenelse{\isempty{#1}}{%
          \texttt{pointed}%
        }{INVALID OPTIONAL}%
      }%
  }{}}}}%
  \texttt{(dim=\textnormal{#3}%
    \ifthenelse{\isempty{#4}}{}{,depth=\textnormal{#4}}%
    \ifthenelse{\isempty{#5}}{}{,\textnormal{#5}}%
  )}%
}

% Non-essential macro
\definecolor{unused}{RGB}{211,167,192}
\newcommand{\unused}[1]{{\color{unused} #1}}

@@

\begin{lem}{}\label{lem:zinf-identities}
  We have the following dual identities for \zinf{} intervals:
%
  \begin{align*}
    \interv{a}{b}{c} \cap \interv{d}{e}{f} & =
      \interv{\max \{a,d\}}{\min \{b,e\}}{c \wedge f} \\
    \interv{a}{b}{c} \cup \interv{d}{e}{f} & =
      \interv{\min \{a,d\}}{\max \{b,e\}}{c \vee f}
  \end{align*}
\end{lem}
%
\begin{proof}
  We give the proof of the first identity and the second one is similar.
  \begin{align*}
    \interv{a}{b}{c} \cap \interv{d}{e}{f} = &
      \; \{ n \;|\; a \leq n \leq b \wedge (\neg c \implies n \neq 0) \}
      \;\cap \\
      & \; \{ n \;|\; d \leq n \leq e \wedge (\neg f \implies n \neq 0) \}
      \\
    = & \; \{ n \;|\; \max \{a,d\} \leq n \leq \min \{b,e\} \wedge (\neg c
      \vee \neg f \implies n \neq 0) \} \\
    = & \; \interv{\max{\{a,d\}}}{\min{\{b,e\}}}{\neg c \vee \neg f}
  \end{align*}
\end{proof}

@@

\unused{
\begin{lem}{}\label{lem:vector-intersect}
  Intersection distributes over offsets vectors.
%
  \begin{equation*}
    (\textit{vint}_1, \dots, \textit{vint}_N) \;\cap\;
      (\textit{vint'}_1, \dots, \textit{vint'}_N)
      =
    (\textit{vint}_1 \;\cap\; \textit{vint'}_1, \dots,
     \textit{vint}_N \;\cap\; \textit{vint'}_N)
  \end{equation*}
\end{lem}
%
\begin{proof}
  TODO. Trivial. Do if needed.
\end{proof}
}

\begin{lem}{}\label{lem:vector-union}
  Let $v$ and $v'$ be two $N$ dimensional offsets vectors such that $v_i$ is
  equal to $v'_i$ for all $1 \leq i \leq N$ except $k$, then we have the
  following:
%
  \begin{equation*}
    (v_1,\dotsc, v_k,\dotsc, v_N) \; \cup \; (v'_1,\dotsc, v'_k, \dotsc, v_N)
    =
    (v_1,\dotsc, v_k \cup v'_k,\dotsc, v_N)
  \end{equation*}
\end{lem}
%
\begin{proof}
  \begin{align*}
    v \; \cup \; v' &
    = \{(x_1,\dotsc,x_n) \mid
          \bigwedge_{1 \leq i \leq N } x_i \in v_i \}
      \;\cup\;
      \{(x_1,\dotsc,x_n) \mid
          \bigwedge_{1 \leq i \leq N } x_i \in v'_i \} \\
    & = \{(x_1,\dotsc,x_n) \mid
          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}} x_i \in v_i
              \wedge x_k \in v_k \vee
          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}} x_i \in v'_i
            \wedge x_k \in v'_k
        \} \\
    & = \{(x_1,\dotsc,x_n) \mid
          \bigwedge_{\substack{1 \leq i \leq N \\ i \neq k}} x_i \in v_i
          \wedge x_k \in v_k \cup v'_k
        \} \\
    & = (v_1,\dotsc, v_k \cup v'_k,\dotsc, v_N)
  \end{align*}
\end{proof}

@@

\unused{
  \begin{lem}{}
    Intersection distributes $N$ dimensional vector intervals:
  %
    \begin{gather*}
      (\interv{a_1}{b_1}{c_1}, \dots, \interv{a_N}{b_N}{c_N}) \; \cap \;
        (\interv{a'_1}{b'_1}{c'_1}, \dots, \interv{a'_N}{b'_N}{c'_N}) \\
      = \\
      (\interv{\max \{a_1,a'_1\}}{\min \{b_1,b'_1\}}{\neg c_1 \vee \neg c'_1},
        \dots, \interv{\max \{a_N,a'_N\}}{\min \{b_N,b'_N\}}{\neg c_N \vee \neg c'_N})
    \end{gather*}
  \end{lem}
  %
  \begin{proof}
    This immediately follows from \Cref{lem:vector-intersect} and
    \Cref{lem:zinf-identities}.
  \end{proof}
}

@@

We give some basic interpretations: $\interp{\texttt{n}} = n$ where $n$ is in
$\mathbb{Z}$, $\interp{\texttt{pointed}} = \llbracket \epsilon \rrbracket =
\textsf{true}$, $\interp{\texttt{nonpointed}} = \textsf{false}$.
Typographically, cursive font is used within specifications to indicate a
partial textual capture. For convenience we use \textit{italic} serif font
inside specifications rather than interpreting its textual capture.  These
coalesces syntactic and semantic representation but avoids clutter in the
presentation.

@@

First, we give meaning to region constants (introduced in \Cref{}).
%
\begin{align*}
  \interp{\stencil{p}{$n$}{}{}}_N & =
    \vecgen{N}{\interp{i}}{\interv{0}{0}{\textsf{true}}}\\
%
  \interp{\stencil{c}{$i$}{$k$}{\textcap{p}}}_N & =
    \vecgen{N}{i}{\interv{-k}{k}{\interp{\textcap{p}}}} \\
%
  \interp{\stencil{f}{$i$}{$k$}{\textcap{p}}}_N & =
    \vecgen{N}{\textcap{i}}{\interv{0}{k}{\interp{\textcap{p}}}} \\
%
  \interp{\stencil{b}{$i$}{$k$}{\textcap{p}}}_N & =
  \vecgen{N}{\textcap{i}}{\interv{-k}{0}{\interp{\textcap{p}}}}
\end{align*}

@@

We know introduce the equational and approximation theory of stencil
specifications based on $\equiv$ and $\preceq$.

@@

\begin{description}
  \item[Basic] \texttt{*} and \texttt{+} are both idempotent, commutative, and
    associative;
%
  \item[Dist] \texttt{*} distributed over \texttt{+} and dually
    \texttt{+} distributes over \texttt{*}, meaning if \textcap{R}, \textcap{S},
    and \textcap{T} are regions, then we have the following equivalences:
%
    \begin{align*}
      \texttt{\textcap{R}*(\textcap{S}+\textcap{T})} & \equiv
        \texttt{(\textcap{R}*\textcap{S})+(\textcap{R}*\textcap{T})} &
      \texttt{\textcap{R}+(\textcap{S}*\textcap{T})} & \equiv
        \texttt{(\textcap{R}+\textcap{S})*(\textcap{R}+\textcap{T})}
    \end{align*}
%
  \item[Overlapping pointed] If \textcap{R} is one of \texttt{forward},
    \texttt{backward}, or \texttt{centered}, then we have
    $\texttt{\textcap{R}(depth=$k$,dim=$n$,\_) + \stencil[s]{p}{$n$}{}{}} \equiv
    \texttt{\textcap{R}(depth=$k$,dim=$n$)}$.
%
  \item[Centered] The region constants \texttt{forward} and \texttt{backward}
    are two halves of \texttt{centered} specifications:
%
    \begin{equation*}
      \stencil[s]{c}{$n$}{$k$}{\textcap{p1}} \equiv
        \stencil[s]{f}{$n$}{$k$}{\textcap{p2}} \texttt{+}
        \stencil[s]{f}{$n$}{$k$}{\textcap{p3}}
    \end{equation*}
%
    Here \textcap{p1} is \texttt{nonpointed} if both \textcap{p2} and
    \textcap{p3} are \texttt{nonpointed} and \texttt{pointed} otherwise.
%
  \item[Point] Let \textcap{R} and \textcap{S} be region constants with two
    variants a superscript \textit{f} to mean nonpointed and \textit{t} to mean
    pointed and \textcap{dr} \& \textcap{ds} be the dimensions \textcap{R} and
    \textcap{S} acts on. Then we have the following dual equivalences:
%
    \begin{align*}
      (\textcap{R}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{f}})
      \;\texttt{+}\;
      (\textcap{R}^{\textit{f}} \texttt{*} \textcap{S}^{\textit{t}})
      \;\texttt{+}\;
      \texttt{p(dim=\textcap{dr})*p(dim=\textcap{ds})} & =
        \textcap{R}^{\textit{t}} \texttt{*} \textcap{S}^{\textit{t}} \\
%
      (\textcap{R}^{\textit{t}} \texttt{+} \textcap{S}^{\textit{f}}) \texttt{*}
      (\textcap{R}^{\textit{f}} \texttt{+} \textcap{S}^{\textit{t}}) & =
        \textcap{R}^{\textit{f}} \texttt{+} \textcap{S}^{\textit{f}}
        \;\texttt{+}\;
        \texttt{p(dim=\textcap{dr})*p(dim=\textcap{ds})}
    \end{align*}
\end{description}

@@

\subsection{Approximations}
We define a partial of order of approximations, $\preceq$. This relation
provides us with the following:

\begin{description}
  \item[Centered]
  \item[Point] Let \textcap{R} be a region constant with two variants: a
    superscript $f$ to mean nonpointed and $t$ to mean pointed. Then we have
    $\textcap{R}^t \preceq \textcap{R}^f$.
\end{description}

@@

\begin{thm}{(Completeness of inference)}
  Inference procedure can infer a specification for every stencil computation.
\end{thm}
%
\begin{proof}
  Would be nice to have.
\end{proof}

\begin{thm}{(Completeness of exact inference)}
  Inference procedure can infer a specification for every stencil computation
  with array indexing behaviour forms a superposition of hyperrectangles removed
  at most one unit from the origin in any of its dimension.
\end{thm}
%
\begin{proof}
  Would be nice to have.
\end{proof}

\begin{thm}{(Optimality of inference)}
  For exact inference, inference procedure always infers specifications with
  fewest number of region constants.
\end{thm}
%
\begin{proof}
  Would be nice to have.
\end{proof}
